---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Let's start by downloading, unzipping and loading the data from the repository.

```{r, cache=TRUE}
if (!file.exists("activity.csv")) {
    download.file(
        "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                  "activity.zip")
    unzip("activity.zip")
}

activity <- read.csv(file = "activity.csv")
```

Check out how the data looks like.

```{r}
head(activity)
str(activity)
```

Copy activity with NA values to **activity.na**:

```{r}
activity.na <- activity
```

Let's filter out the NA values.

```{r}
activity <- subset(activity, !is.na(steps))
```

Check out how the filtered data looks like.

```{r}
head(activity)
str(activity)
```

## What is mean total number of steps taken per day?

Total number of steps for each days:

```{r}
steps.per.day <- tapply(activity$steps, activity$date, sum)
steps.per.day[is.na(steps.per.day)] <- 0
steps.per.day
```

Now visualize it with a histogram:

```{r}
barplot(steps.per.day, col = "red")
```

The mean and the median of the total steps per day (different than 
***mean/median of steps per day***)

```{r}
mean(steps.per.day)
median(steps.per.day)
```


> **NOTE: ** Results may vary because total steps with NA per day are taken as 0
for continuous time series.

## What is the average daily activity pattern?

First of all take means of steps for each interval across all days.

```{r}
steps.per.interval <- tapply(activity$steps, activity$interval, mean)
steps.per.interval
```

After that get levels of intervals as an array called **Interval**.

```{r}
Interval <- as.numeric(levels(as.factor(activity$interval)))
Interval
```

Finally plot the time series using **base** plot system.

```{r}
plot(Interval, steps.per.interval, type = "l", ylab = "Number of steps")
```

Now it's easy to retrieve the interval with the maximum steps:

```{r}
max.steps <- max(steps.per.interval)
max.interval <- steps.per.interval[steps.per.interval == max.steps]
as.numeric(names(max.interval))
```

The interval with the maximum steps is **835** (i.e. 8:35-8:40)

## Imputing missing values

Let's start by counting rows with *NA* values:

```{r}
sum(is.na(activity.na$steps))
```

In order to impute we use the median strategy mixed with interpolation. 
1. If there is a missing value between two existing values take the mean of both values
2. Otherwise impute with that hour's mean

```{r}
activity.imputed <- activity.na
activity.imputed$steps[1] <- steps.per.interval[1]
length.activity <- nrow(activity.imputed)
for (i in 2:length.activity - 1) {
    if (is.na(activity.imputed$steps[i])) {
        prev <- activity.imputed$steps[i - 1]; nxt <- activity.imputed$steps[i + 1]
        activity.imputed$steps[i] <- if (!is.na(prev) && !is.na(nxt)) {
            (prev + nxt) / 2
        } else {
            steps.per.interval[as.numeric(names(steps.per.interval)) == 
                                   activity.imputed$interval[i]]
        }
    }
}
activity.imputed$steps[length.activity] <- steps.per.interval[
    as.numeric(names(steps.per.interval)) == activity.imputed$interval[length.activity]]
```

Now check out the new data:

```{r}
head(activity.imputed)
tail(activity.imputed)
str(activity.imputed)
```

Check out how does it look like in comparison with the original dataset:

```{r}
steps.per.day.imputed <- tapply(activity.imputed$steps, activity.imputed$date, sum)
steps.per.day.imputed[is.na(steps.per.day)] <- 0
par(mfrow = c(2, 1))
barplot(steps.per.day, col = "red", main = "Without Imputing")
barplot(steps.per.day.imputed, col = "blue", main = "Imputed")
```

Now calculate the new mean and median and compare it to the original:

```{r results="asis"}
library(xtable)
dt.original <- data.frame(mean = mean(steps.per.day), 
                          median = median(steps.per.day))
dt.imputed <- data.frame(mean = mean(steps.per.day.imputed), 
                         median = median(steps.per.day.imputed))
dt.table <- rbind(dt.original, dt.imputed)
row.names(dt.table) <- c("original", "imputed")
xt <- xtable(dt.table)
print(xt, type = "html")
```


This table clearly shows that imputing increased both median and mean and 
also made mean and median closer to each other because of eliminating sudden 
jumps in data.


## Are there differences in activity patterns between weekdays and weekends?
